How to select predictive models for decision-making or causal inference?

### 2023-12-08 how_to_select_causal_models.tex
iter

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.%TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchors causal model selection in the \emph{potential
    outcome} framework and details the causal risks and model-selection procedure.
We then rewrite the so-called $R\text{-risk}$, as a reweighted version of mean
squared difference between the true and estimated individualized treatment
effect. Finally, we conduct a thorough empirical study comparing the different
metrics on a wide variety of data-generative processes based on simulations and
real data.

@@

    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
                models} are good overall estimators of the nuisances}:
        Results are shown only for the
        R-risk; Figure \ref{apd:fig:nuisances_comparison}
        details every metrics. For Twins, where the true propensity
        model is linear, \textcolor{DarkOrchid}{stacked} and
        \textcolor{ForestGreen}{linear}
        estimations of the nuisances performs equivalently, even for a downsampled version
        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
\end{figure}

@@

\section{Discussion and conclusion}\label{sec:discussion}

Predictive models are increasingly used to reason about causal effects,
for instance in precision medicine to drive individualized decision.
Our results highlight that they should be selected, validated, and tuned
using different procedures and error measures than those classically used
to assess prediction (estimating the so-called $\mu\text{-risk}$).
Rather, selecting the best outcome model according to the $R\text{-risk}$
(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.

\paragraph{Nuisance models: more gain than pain}
%
Estimating the $R\text{-risk}$ requires a more complex procedure than
standard cross-validation used \emph{e.g.}~in machine learning: it involves
fitting nuisance models necessary for model evaluation, though our
results show that these can be learned on the same set of data
as the outcome model evaluated.

The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). However these models are
easier to select and control than a causally-valid outcome model,
as they are associated to errors on
observed distributions. Our results show that using for nuisance models
a flexible stacking-based family of estimator suffices for good model selection.
%
In fact, a feasible $R\text{-risk}$ --where the nuisances
are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
nuisances are known. This may be explained by
results that suggest that estimation errors on both
nuisances partly compensate out in the
$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.

Note that propensity score models must be selected to estimate
the individual posterior probability. For this, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\paragraph{More $R\text{-risk}$ to select models driving decisions}

>>>>

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

We first illustrate the problem of causal model selection and briefly review
prior art. Then we anchor causal model selection in the \emph{potential
    outcome} framework and details the causal risks and model-selection procedure.
We then rewrite the so-called $R\text{-risk}$, as a reweighted version of mean
squared difference between the true and estimated individualized treatment
effect. Finally, we conduct a thorough empirical study comparing the different
metrics on a wide variety of data-generative processes based on simulations and
real data.

@@

    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
                models} are good overall estimators of the nuisances}:
        Results are shown only for the
        R-risk; Figure \ref{apd:fig:nuisances_comparison}
        details every metrics. For Twins, where the true propensity
        model is linear, \textcolor{DarkOrchid}{stacked} and
        \textcolor{ForestGreen}{linear}
        estimations of the nuisances performs equivalently, even for a downsampled version
        (N=4,794). }\label{fig:all_datasets_nuisances_comparison}
\end{figure}

@@

\section{Discussion}\label{sec:discussion}

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about causal effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction
(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \emph{e.g.}~in machine
learning: it involves fitting nuisance models necessary for model evaluation,
though our results show that these can be learned on the same set of data as the
outcome model evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. For propensity score, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

%
% However these models are easier to select and control than a causally-valid
% outcome model, as they are associated to errors on observed distributions. 
% In fact, a feasible $R\text{-risk}$ --where the nuisances
% are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
% nuisances are known. This may be explained by
% results that suggest that estimation errors on both
% nuisances partly compensate out in the
% $R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.

@@

\section{Conclusion}\label{sec:conclusion}
\paragraph{More $R\text{-risk}$ to select models driving decisions}
